Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-10-06 article.Rmd
Adding more context, citations

This paper addresses two trends in the humanities: the rise of gender as an analytical tool and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name using historical data. In contrast to existing methods of predicting gender, this method takes into account how naming practices change over time and space. It provides a higher level of accuracy and a precise statement of the likelihood that a name was a given gender, especially for datasets where names can be associated with dates. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we study gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*.

### Introduction: The "Leslie" Problem

@@

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, existing tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie" problem is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. As of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between their births and deaths [World Bank 2014]. Predicting gender from first names therefore requires a fundamentally historical method.

@@

The Social Security Administration (SSA) "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 [Social Security Administration 2014-b].] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security became available during the New Deal, its first beneficiaries were adults past or near retirement age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studied the history of birth registration in the United States [Landrum 2014]. It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. These coins are often measured takes the form of book reviews published in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the *AHR* publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." The *AHR* is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, and longevity of the *AHR* makes it an ideal source to analyze the historical profession on a large scale.

We take the <i>American Historical Review</i> as a rough proxy for gatekeeping within the larger historical profession. When a book appears in the *AHR* it serves as a signal that other historians in positions of power have "approved" it for consideration. This signal is not necessarily correlated with the quality of the book. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review. To be fair, a book that does not appear in the *AHR* may have gone on to have a significant impact on the profession. It is precisely this professional gatekeeping dimension that makes the *AHR* useful for studying gender representation.

Scraping the table-of-contents of every issue of the American Historical Review results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could then use to plot the proportion of female and male authors over time. The temporal trajectory of gender representations roughly resembles that of history dissertations: women began making inroads in the late 1960s and have made steady gains over the past four decades. By the twenty-first century the proportion of female authors reviewed in the *AHR* had more than tripled.

@@

The path from a dissertation to a monograph to a book review in the *American Historical Association* isn't automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troubling, women continue to face gender discrimination in the professional sphere.

We used our program to infer the gender of not just book authors reviewed by the *AHR*, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

@@

@@

[Weingart 2014] Weingart, Scott. "The Moral Role of DH in a Data-Driven World." *the scottbot irregular*. 9 September 2014, http://www.scottbot.net/HIAL/?p=40944

>>>>

This paper addresses two trends in the humanities: the use of gender as an analytical framework and the proliferation of large, machine-readable datasets. It outlines a new method for algorithmically inferring the gender of a name using historical data. In contrast to existing methods of predicting gender, this method takes into account how naming practices change over time and space. It provides a higher level of accuracy and a precise statement of the likelihood that a name was a given gender, especially for datasets where names can be associated with dates. The article describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we study gender and gatekeeping in the history profession by inferring the gender of authors of history dissertations and of authors and reviewers in the *American Historical Review*. The Gender package illustrates how humanities scholars can make concrete methodologial contributions to computer science and related fields.

### Introduction: The "Leslie Problem"

@@

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same education researcher wanted to compare the demographics of contemporary urban schools with those from the early twentieth century, existing tools like Genderize.io would misidentify Leslie Fay's gender in historical datasets. The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. As of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between their births and deaths [World Bank 2014]. Predicting gender from first names therefore requires a fundamentally historical method.

@@

The Social Security Administration (SSA) [Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data) dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 [Social Security Administration 2014-b].] This dataset contains a list of first names along with how many times each name was assigned to each gender in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security became available during the New Deal, its first beneficiaries were adults past or near retirement age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studied the history of birth registration in the United States [Landrum 2014]. It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

But what happens after the dissertation? It is, after all, only the first major stage of a historian’s research output. The "coin of the realm" for many historians remains the publication of a scholarly monograph to be read and evaluated by peers. Once published, the impact of a book is often shaped by its reviews in academic journals. One of the leading journals in the historical profession is the *American Historical Review*, the flagship journal of the American Historical Association. Along with a few dozen articles published in its five issues each year, the *AHR* publishes roughly one thousand books reviews per year covering (in its own words) "every major field of historical study." The *AHR* is not only one of the widest-ranging journals in the profession, it is also the oldest; the journal has been publishing continuously since 1895. The range, scope, longevity, and reputation of the *AHR* makes it an ideal proxy to study gatekeeping within the larger historical profession.

A few caveats are in order. First, a book that never appears in the *AHR* at all may still go on to have a significant impact on the profession. Second, a book's appearance in the *AHR* is not necessarily correlated with the *quality* of the book, as the journal prints negative reviews as well as positive ones. Nevertheless, when a book appears in the *AHR* it serves as a signal that other historians in positions of power have "approved" it for consideration. Even if it garners a negative review, the fact that it appears in the journal at all is a measure of the fact that the profession's gatekeepers have deemed it important enough to review.  It is precisely this professional gatekeeping dimension that we can analyze using the Gender package.

Scraping the table-of-contents of every issue of the American Historical Review results in a dataset of close to 60,000 books reviewed by the journal since it began publication in 1895. Our program then inferred the gender of the authors of these books, which we could in turn use to plot the proportion of female and male authors over time. The temporal trajectory of gender representations roughly resembles that of history dissertations: women began making inroads in the late 1960s and have made steady gains over the past four decades. By the twenty-first century the proportion of female authors reviewed in the *AHR* had more than tripled.

@@

The path from a dissertation to a monograph to a book review in the *American Historical Association* isn't automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2011 survey by the American Historical Association revealed that female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. [Townsend 2012] This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. Their male colleagues, meanwhile, enjoy more time to write and receive more accolades for their writing. Since 2004, the Pulitzer Prize for History has been awarded to just one female historian ["The Pulitzer Prizes: History" 2014-], while the Bancroft Prize in American History has been awarded to 24 men and 6 women over that same period ["Bancroft Prize"]. The *American Historical Review* is, unfortunately, far from alone in its gatekeeping practices.

We went on to use our program to infer the gender of not just book authors reviewed by the *AHR*, but the gender of the reviewers themselves. The story is much the same for reviewers as it is for authors: more than twice as many men as women appear as reviewers in the journal. But gender inflects book reviews in less direct ways by shaping who writes reviews of which book authors. About three times as many men write reviews of male-authored books as do women. In the case of female-authored books the ratio of female-to-male reviewers is much closer to 50/50. In short, women are much more likely to write reviews of other women. And while men still write the majority of reviews of female-authored books, they tend to gravitate more towards male authors – who are, of course, already over-represented in the *AHR*.

@@

["Bancroft Prize"] "Bancroft Prize." *Wikipedia* <http://en.wikipedia.org/wiki/Bancroft_Prize> Accessed 5 October 2014.

@@

["The Pulitzer Prizes: History" 2014-] "The Pulitzer Prizes: History." 2014-. <http://www.pulitzer.org/bycat/History> Accessed 5 October 2014.

@@

[Townsend 2012] Townsend, Robert B. "What Makes a Successful Academic Career in History? A Field Report from the Higher Ranks." *Perspectives on History*. December 2012, <http://www.historians.org/publications-and-directories/perspectives-on-history/december-2012/what-makes-a-successful-academic-career-in-history>

[Weingart 2014] Weingart, Scott. "The Moral Role of DH in a Data-Driven World." *the scottbot irregular*. 9 September 2014, <http://www.scottbot.net/HIAL/?p=40944>
